---
title: "Ancestral state reconstruction explore"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Using phytools and ape for ancestral state reconstruction

Refer http://www.phytools.org/eqg2015/asr.html

```{r}
library(phytools)
library(ape)
```

## Read the tree

```{r}
#tree <- read.tree("https://raw.githubusercontent.com/Ensembl/ensembl-compara/release/94/scripts/pipeline/species_tree.ensembl.branch_len.nw")
#tree <- read.tree("http://hgdownload.cse.ucsc.edu/goldenpath/hg38/multiz17way/hg38.17way.nh")
#tree$tip.label <- sub("[0-9]*$", "", tree$tip.label)
#tree <- keep.tip(tree, c("hg", "gorGor", "panTro", "ponAbe", "rheMac"))
tree <- read.tree(text = "((((hg:6.4[&&NHX:dist=6.4:name=hg:support=1.0],panTro:6.4[&&NHX:dist=6.4:name=panTro:support=1.0])1:2.21[&&NHX:dist=2.21:name=:support=1.0],gorGor:8.61[&&NHX:dist=8.61:name=gorGor:support=1.0])1:6.59[&&NHX:dist=6.59:name=:support=1.0],ponAbe:15.2[&&NHX:dist=15.2:name=ponAbe:support=1.0])1:12.9[&&NHX:dist=12.9:name=:support=1.0],rheMac:28.1[&&NHX:dist=28.1:name=rheMac:support=1.0]);")
tree
plotTree(tree)
```

## Load sample SNP data

```{r}
library(here)
snps <- local({
    snps <- readr::read_tsv(here("results/snptable_2018-11-20_04:15/1523_2018-11-20_04:18.txt.gz"))
    
    ## Only select some SNPs
    idx <- apply(snps[! (names(snps) %in% c("ALT", "Loc"))], 1, function(r) {
        length(unique(r)) >= 4
    })
    snps <- snps[idx, ]
    
    # Fill with NA
    for (i in seq_along(snps)) {
        col <- snps[[i]]
        col[col == "."] <- NA
        snps[[i]] <- col
    }
    
    snps
})
snps
```

Function to convert the rows to tip list.

```{r}
extract_tips <- function(snps, idx = 42) {
    stopifnot(length(idx) == 1)
    tips <- snps[idx, ]
    tips$Loc <- NULL
    tips$ALT <- NULL
    names(tips)[names(tips) == "REF"] <- "hg38"
    names(tips) <- sub("[0-9]*$", "", names(tips))
    unlist(as.list(tips))
}
```


## Ancestral character estimation

We use maximum likelihood estimation for the discrete characters.
For the transition/transversion rate matrix, we use a Kimura two parameter (K2P) model
with a default ratio of transitions to transversions of 1.5 (which may be adjusted later).

```{r}
K2PModel <- function(R = 1.5, N = c("A", "T", "C", "G")) {
    # R is the the expected ratio of transitions to transversions.
    # Refer http://www.cbs.dtu.dk/courses/27615/exercises/distance2.php.save
    # FIXME: choose another default R value?
    b <- 1
    a <- 2 * R * b
    m <- matrix(c(0, b, a, b,
                  b, 0, b, a,
                  a, b, 0, b,
                  b, a, b, 0), byrow = TRUE, ncol = 4)
    colnames(m) <- c("A", "C", "G", "T")
    rownames(m) <- c("A", "C", "G", "T")
    N <- sort(na.omit(unique(N)))
    m <- m[N, N, drop = FALSE]
    
    # TMP
    #m <- m[, rev(1:ncol(m))]
    
    m
}
```

The default matrix:

```{r}
K2PModel()
```

### A sample prediction

We just try one pair of values first.

```{r}
tips <- extract_tips(snps, 88)
tips
fit <- ace(tips, tree, type = "discrete", model = K2PModel(N = tips), marginal = FALSE)
fit
```

The element `lik.anc` gives the scaled likelihoods of each ancestral states at each node.

```{r}
round(fit$lik.anc, 3)
```


### Visualize the predicted ancestral state

```{r}
vis_ace <- function(fit, tree, tips) {
    
    ## A, C, T, G + "Black" for NA state
    cols <- RColorBrewer::brewer.pal(4, "Set3")
    cols <- c(cols, "black")
    cols_order <- c("A", "C", "G", "T", "?")
    
    ## Plot Tree
    mod_tip_labels <- function(tree, tips) {
        ### Modify the tip labels
        tips <- tips[tree$tip.label] ## Put tips in the same order
        tips[is.na(tips)] <- "?"
        tree$tip.label <- paste0(tree$tip.label, " (", tips, ")")
        tree
    }
    plotTree(mod_tip_labels(tree, tips), offset = 1)
    
    ## Plot node labels
    norm_nodepie <- function(pie) {
        ## Fill extra columns and order the columns in order to
        ## make the color scale constant.
        for (coln in cols_order) {
            if (! coln %in% colnames(pie))
                pie <- cbind(pie, matrix(0, nrow(pie), dimnames = list(NULL, coln)))
        }
        pie <- pie[, cols_order]
        pie
    }
    pie.node <- norm_nodepie(fit$lik.anc)
    nodelabels(
        node = Ntip(tree) + 1:Nnode(tree),
        pie = pie.node,
        piecol = cols,
        cex = 1
    )
    
    ## Plot tip labels
    tips <- tips[tree$tip.label] ## Put tips in the same order
    pie.tips <- to.matrix(tips, cols_order)
    pie.tips[, "?"] <- apply(pie.tips, 1, function(r) if (sum(r) == 0) 1 else 0)
    tiplabels(
        pie = pie.tips,
        piecol = cols,
        cex = 0.8
    )
    
    ## Add legend
    add.simmap.legend(
        colors = setNames(cols, cols_order),
        x = 0, y = 2, prompt = FALSE
    )
}
vis_ace(fit, tree, tips)
```


## Check some values

```{r eval=FALSE, include=FALSE}
acesnp <- function(which) {
    tips <- extract_tips(snps, which)
    fit <- ace(tips, tree, type = "discrete", model = K2PModel(N = tips), marginal = FALSE)
    vis_ace(fit, tree, tips)
}
for (i in 1:100) {
    #Sys.sleep(1)
    print(i)
    acesnp(i)
}
```

```{r}
acesnp2 <- function(hg = "A", panTro = "A", gorGor = "A", ponAbe = "A", rheMac = "A", 
                    marginal = FALSE) {
    tips <- c(hg = hg, panTro = panTro, gorGor = gorGor, ponAbe = ponAbe, rheMac = rheMac)
    fit <- ace(tips, tree, type = "discrete", model = K2PModel(N = tips),
               marginal = marginal)
    
    #print(K2PModel(N = tips))
    #print(fit$index.matrix)
    #print(fit$lik.anc)
    #print(fit$rates)
    
    vis_ace(fit, tree, tips)
    invisible(fit)
}
acesnp2("A", "A", "C", "A", "A")
acesnp2("A", "A", "C", "C", "C")
acesnp2("A", "C", "C", "C", "C")

acesnp2("A", "A", "A", "A", "C")
acesnp2("A", "A", "A", "C", "C")
acesnp2("A", "A", "C", "C", "C")
acesnp2("A", "C", "C", "C", "C")

acesnp2("A", "A", "A", "A", "T")
acesnp2("A", "A", "A", "A", "G")
acesnp2("A", "A", "A", "A", "C")

acesnp2("A", "A", "A", "C", "T")
acesnp2("A", "A", "A", "C", "G")
acesnp2("A", "A", "A", "C", "C")

acesnp2("A", "A", "A", "C", "G", marginal = FALSE)
acesnp2("A", "A", "A", "C", "G", marginal = TRUE)
```

