---
title: "Try phangorn"
output:
  html_document:
    toc: yes
    self_contained: false
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Phangorn

I have tried `ace` function from `ape` package which uses maximum likelihood method.
But I think the ML method may not work well when the tree is small.
The `MPR` function is an alternative method which uses most parsimonious reconstruction.

Here I would like to try the `phangorn` package. It has both parsimony method and maximum
likelihood method to predict the ancestral state.

Reference: https://cran.r-project.org/web/packages/phangorn/vignettes/Ancestral.pdf

```{r}
library(phangorn)
```


First to read the tree.

```{r}
tree <- read.tree(text = "((((hg:6.4[&&NHX:dist=6.4:name=hg:support=1.0],panTro:6.4[&&NHX:dist=6.4:name=panTro:support=1.0])1:2.21[&&NHX:dist=2.21:name=:support=1.0],gorGor:8.61[&&NHX:dist=8.61:name=gorGor:support=1.0])1:6.59[&&NHX:dist=6.59:name=:support=1.0],ponAbe:15.2[&&NHX:dist=15.2:name=ponAbe:support=1.0])1:12.9[&&NHX:dist=12.9:name=:support=1.0],rheMac:28.1[&&NHX:dist=28.1:name=rheMac:support=1.0]);")
tree$node.label <- c("A", "B", "C", "D")
tree
```

## Predict some cases

```{r}
fit_anc <- function(tree, tips, ...) {
    stopifnot(is.list(tips))
    if (is.null(names(tips)))
        names(tips) <- tree$tip.label
    stopifnot(identical(names(tips), tree$tip.label))
    tips <- phyDat(tips, type = "DNA")
    cat(" - parsimony score:", parsimony(tree, tips), "\n")
    ans <- ancestral.pars(tree, tips, ...)
    #print(ans) # TODO: make sure sitePattern == 1
    ans
}
```

When there are two alternative alleles in human:

```{r, out.width="60%", fig.dim=c(7,5)}
plotAnc(tree, fit_anc(tree, list("V", "C", "G", "G", "G")))
plotAnc(tree, fit_anc(tree, list("V", "C", "A", "G", "G")))
plotAnc(tree, fit_anc(tree, list("V", "C", "A", "A", "G")))
plotAnc(tree, fit_anc(tree, list("V", "C", "A", "A", "C")))
plotAnc(tree, fit_anc(tree, list("V", "C", "C", "A", "C")))
plotAnc(tree, fit_anc(tree, list("V", "C", "?", "A", "C")))
plotAnc(tree, fit_anc(tree, list("V", "C", "?", "A", "?")))
plotAnc(tree, fit_anc(tree, list("V", "C", "?", "A", "G"))) # In this case, we know nothing..
```

When there are two states in human.

```{r, out.width="60%", fig.dim=c(7,5)}
plotAnc(tree, fit_anc(tree, list("W", "A", "A", "T", "G")))
plotAnc(tree, fit_anc(tree, list("W", "A", "A", "T", "A")))
plotAnc(tree, fit_anc(tree, list("W", "A", "?", "T", "A")))
plotAnc(tree, fit_anc(tree, list("W", "A", "?", "T", "?")))
plotAnc(tree, fit_anc(tree, list("W", "A", "?", "A", "?")))
plotAnc(tree, fit_anc(tree, list("W", "A", "T", "A", "G")))
plotAnc(tree, fit_anc(tree, list("W", "A", "T", "T", "G")))
```

## Add cost matrix

```{r}
COST <- function(ratio) {
    levels <- c("a", "c", "g", "t")
    transversion_cost <- ratio
    cost <- matrix(transversion_cost, 4, 4, dimnames = list(levels, levels))
    cost <- cost - transversion_cost * diag(4)
    cost
    transitions <- rbind(
        c("a", "g"),
        c("g", "a"),
        c("c", "t"),
        c("t", "c")
    )
    cost[transitions] <- 1
    unname(cost)
}
fit_anc2 <- function(tree, tips, cost = COST(1.1)) {
    stopifnot(is.list(tips))
    if (is.null(names(tips)))
        names(tips) <- tree$tip.label
    stopifnot(identical(names(tips), tree$tip.label))
    tips <- phyDat(tips, type = "DNA")
    ans <- ancestral.pars(tree, tips, return = "prob", cost = cost)
    #tmp <<- ancestral.pars(tree, tips, return = 'phyDat')
    ans
}
```

```{r}
# A to G (transition) > A to T (transversion).
plotAnc(tree, fit_anc2(tree, list("G", "T", "A", "A", "A")))
plotAnc(tree, fit_anc2(tree, list("T", "C", "A", "A", "A")))

# Another case
plotAnc(tree, fit_anc2(tree, list("T", "T", "G", "A", "A")))

# Does this make sense ???
plotAnc(tree, fit_anc2(tree, list("C", "A", "A", "A", "T"), cost = COST(9)))
plotAnc(tree, fit_anc2(tree, list("C", "A", "A", "A", "T"), cost = COST(1.1)))
```

## Function for wrapping multiple predictions

```{r}
fit_anc3 <- function(tree, phyd) {
    #' @phyd: A phyDat object
    stopifnot(identical(names(phyd), tree$tip.label))
    cost <- COST(1.1)
    ans <- ancestral.pars(tree, phyd, return = "phyDat", cost = cost)
    ans
}
```


## TODO:

1. a cost matrix (Not transitions rates!!) for the transitions between two states
    may be specified to the function. How to convert K2P Model to cost matrix?

2. Try `ancestral.pars(type = "ACCTRAN")`, the results can be slightly different, what's
    the difference?

